Follow-up Samples

### 2017-09-15 submission/manuscript_R2.Rmd
add missing parenthesis

**Results.** There were small changes to the bacterial community associated with adenoma or advanced adenoma and large changes associated with carcinoma. The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`). Although treatment was associated with intrapersonal changes, the change in the abundance of individual OTUs in response to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05). Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to identify groups of OTUs that could be used to classify pre and post-treatment samples for each of the diagnosis groups. Although the three models could differentiate between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of each treatment. To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas. We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treatment samples. Samples from patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.001, invisible("<"), invisible(">"))` 0.001). Finally, we were unable to detect any significant differences in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).

@@

***Study Design and Patient Sampling.***
Sampling and design have been previously reported in Baxter, et al [@baxter_microbiota-based_2016]. Briefly, samples were stored on ice for at least 24h before freezing. Although we cannot exclude that this sampling protocol may have impacted the gut microbiota composition all samples were subjected to the same methodology. Study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis. Samples used to build the models for prediction were collected either prior to a colonoscopy or between one and two weeks after initial colonoscopy. The bacterial community has been shown to normalize back to a pre-colonoscopy community within this time period [@obrien_impact_2013]. Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion [Table 1]. Patients were diagnosed by colonoscopic examination and histopathological review of any biopsies taken. Patients were classified as having advanced adenoma if they had an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology. This study was approved by the University of Michigan Institutional Review Board. All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

>>>>

**Results.** There were small changes to the bacterial community associated with adenoma or advanced adenoma and large changes associated with carcinoma. The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`). Although treatment was associated with intrapersonal changes, the change in the abundance of individual OTUs in response to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05). Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to identify groups of OTUs that could be used to classify pre and post-treatment samples for each of the diagnosis groups. Although the adenoma and carcinoma models could reliably differentiate between the pre and post-treatment samples (P-value treatment models had AUCs that were significantly higher than a random model permutation (P-value `r ifelse(filter(treat_random_pvalues, study == "adn")[, "pvalue"] > 0.0001 & filter(treat_random_pvalues, study == "crc")[, "pvalue"] > 0.0001, invisible(">"), invisible("<"))` 0.001)), the advanced-adenoma model could not (P-value = `r format(filter(treat_random_pvalues, study == "srn")[, "pvalue"], digits = 2, nsmall = 2)`). Furthermore, there was little overlap between the OTUs that were indicative of each treatment. To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas. We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treatment samples. Samples from patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.001, invisible("<"), invisible(">"))` 0.001). Finally, we were unable to detect any significant differences in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).

@@

***Study Design and Patient Sampling.***
Sampling and design have been previously reported in Baxter, et al [@baxter_microbiota-based_2016]. Briefly, samples were stored on ice for at least 24h before freezing. Although we cannot exclude that this sampling protocol may have impacted the gut microbiota composition all samples were subjected to the same methodology. Study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis. Samples used to build the models for prediction were collected either prior to a colonoscopy or between one and two weeks after initial colonoscopy. The bacterial community has been shown to normalize back to a pre-colonoscopy community within this time period [@obrien_impact_2013]. Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion [Table 1]. Patients were diagnosed by colonoscopic examination and histopathological review of any biopsies taken. Patients were classified as having advanced adenoma if they had an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology. This study was approved by the University of Michigan Institutional Review Board. All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration. The original protocol for the study did not provide for tracking patients after the follow up samples and so it was not possible for us to ascertain their diagnosis after the completion of the study.

### 2017-09-15 submission/manuscript_R2.tex
add missing parenthesis

\textbf{Results.} There were small changes to the bacterial community
associated with adenoma or advanced adenoma and large changes associated
with carcinoma. The communities from patients with carcinomas changed
significantly more than those with adenoma following treatment (P-value
\textless{} 0.001). Although treatment was associated with intrapersonal
changes, the change in the abundance of individual OTUs in response to
treatment was not consistent within diagnosis groups (P-value
\textgreater{} 0.05). Because the distribution of OTUs across patients
and diagnosis groups was irregular, we used the Random Forest machine
learning algorithm to identify groups of OTUs that could be used to
classify pre and post-treatment samples for each of the diagnosis
groups. Although the adenoma and carcinoma models could reliably
differentiate between the pre and post-treatment samples (P-value
treatment models had AUCs that were significantly higher than a random
model permutation (P-value \textless{} 0.001), the advanced-adenoma
model could not (P-value = 0.61). Furthermore, there was little overlap
between the OTUs that were indicative of each treatment. To determine
whether individuals who underwent treatment were more likely to have
OTUs associated with normal colons we used a larger cohort that
contained individuals with normal colons and those with adenomas,
advanced adenomas, and carcinomas. We again built Random Forest models
and measured the change in the positive probability of having one of the
three diagnoses to assess whether the post-treatment samples received
the same classification as the pre-treatment samples. Samples from
patients who had carcinomas changed towards a microbial milieu that
resembles the normal colon after treatment (P-value \textless{} 0.001).
Finally, we were unable to detect any significant differences in the
microbiota of individuals treated with surgery alone and those treated
with chemotherapy or chemotherapy and radiation (P-value \textgreater{}
0.05).

>>>>

\textbf{Results.} There were small changes to the bacterial community
associated with adenoma or advanced adenoma and large changes associated
with carcinoma. The communities from patients with carcinomas changed
significantly more than those with adenoma following treatment (P-value
\textless{} 0.001). Although treatment was associated with intrapersonal
changes, the change in the abundance of individual OTUs in response to
treatment was not consistent within diagnosis groups (P-value
\textgreater{} 0.05). Because the distribution of OTUs across patients
and diagnosis groups was irregular, we used the Random Forest machine
learning algorithm to identify groups of OTUs that could be used to
classify pre and post-treatment samples for each of the diagnosis
groups. Although the adenoma and carcinoma models could reliably
differentiate between the pre and post-treatment samples (P-value
treatment models had AUCs that were significantly higher than a random
model permutation (P-value \textless{} 0.001)), the advanced-adenoma
model could not (P-value = 0.61). Furthermore, there was little overlap
between the OTUs that were indicative of each treatment. To determine
whether individuals who underwent treatment were more likely to have
OTUs associated with normal colons we used a larger cohort that
contained individuals with normal colons and those with adenomas,
advanced adenomas, and carcinomas. We again built Random Forest models
and measured the change in the positive probability of having one of the
three diagnoses to assess whether the post-treatment samples received
the same classification as the pre-treatment samples. Samples from
patients who had carcinomas changed towards a microbial milieu that
resembles the normal colon after treatment (P-value \textless{} 0.001).
Finally, we were unable to detect any significant differences in the
microbiota of individuals treated with surgery alone and those treated
with chemotherapy or chemotherapy and radiation (P-value \textgreater{}
0.05).
